\documentclass{article}

\usepackage[margin=1in]{geometry}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{tikz}
\usetikzlibrary{automata,positioning}


\pagestyle{fancy}

\lhead{CS 373 \\ Homework 4}
\chead{}
\rhead{Drew Cross (ddcross2) \\
       \emph{Partners:} Eric Parsons}
\lfoot{}
\cfoot{}
\rfoot{}

\begin{document}

\subsection*{Problem 1}
CFG Parsing

RNA molecules are crucial to all life on Earth.  In fact, they're so important, some
researchers theorize these molecules may have been the start of all life on Earth. For our
purposes, RNAs are represented as strings on the alphabet $\Sigma=\{a,u,c,g\}$.  The following 
grammar comes from an actual research pape
\footnote{http://www.biomedcentral.com/1471-2105/5/71} about computational analysis of RNA: \\

    $S \rightarrow LS \;|\; L$

    $L \rightarrow aFu \;|\; uFa \;|\; cFg \;|\; gFc \;|\; a \;|\; u \;|\; c \;|\; g$

    $F \rightarrow aFu \;|\; uFa \;|\; cFg \;|\; gFc \;|\; LS$

\begin{enumerate}[(a)]
\item Put this grammar in CNF. (3 Points)
%TODO
\item Provite the CYK table for these strings:
    \begin{itemize}
        \item $w_1=gaccuc$ (4 points)

        %TODO
        \begin{tabular}{|c|c|c|c|c|c|c|}
        \hline
        & & & & & \\
        \hline
        & & & & & \\
        \hline
        & & & & & \\
        \hline
        & & & & & \\
        \hline
        & & & & & \\
        \hline
        & & & & & \\
        \hline
        g&a&c&c&u&c\\
        \hline
        \end{tabular}
        \item $w_2=uggu$ (3 points)

        %TODO
        \begin{tabular}{|c|c|c|c|}
        \hline
        & & & \\
        \hline
        & & & \\
        \hline
        & & & \\
        \hline
        & & & \\
        \hline
        u&g&g&u\\
        \hline
        \end{tabular}

    \end{itemize}

\item Prove that the language accepted by this grammar is actually regular. (3 Points)

This language is regular because it is essentaily $\Sigma^*$.
%TODO

\item Can you explain why these researchers are interested in parsing a regular language with
a CFG? (2 points)
%TODO
\end{enumerate}

\newpage
\subsection*{Problem 2}
Provide two proofs that the following languages are not context-free. First, provide a proof
using the the pumping lemma for CFLs. Second, provide a proof using closure properties of
CFLs. You may assume that the languages Sipser used when demonstrating the pumping lemma
(see Sipser Examples 2.36, 2.37, and 2.38 pp. 126) are known to be non-context-free.

\begin{enumerate}[(a)]
    \item $\{ 0^n 1^{2m} 0^{m+k} \;|\; 0 \le n \le m, k \ge 0 \}$
    \item $\{w\in \{a,b,c\}^* \;|\; \#a(w)=\#b(w)=\#c(w)\}$
\end{enumerate}

\newpage
\subsection*{Problem 3}
In class, we’ve covered several classes in the Chomsky Hierarchy: regular, context-free, and
context-sensitive languages. Based on what you’ve learned, it should be possible to assign
languages from these classes to their most exclusive (or least general) class in the Chomsky
Hierarchy. For example, we know the language L = f0 n 1n 0g is context-free, but not regular.
We used the pumping lemma for regular languages to prove that it is not regular (see Sipser,
Example 1.73, pp. 80) and wrote a CFG to prove that it is context-free (see Sipser, Section
2.1, pp. 100). Using all your knowledge of formal languages, place these languages into the
most exclusive class in the hierarchy you know. State the class, and provide a proof that it is
a member of that class. Also provide a proof that it is not a member of a more exclusive
class. None of these languages belong exclusively to classes we haven’t yet covered, and all
of the languages are on the alphabet $\Sigma = \{0,1\}$.
\begin{enumerate}[(a)]
    \item $\{w \;|\; \#0(w)=\#1(w)\}$ (5 points)
    \item $\{w \;|\; w=w^r, \#0(w)=\#1(w)\}$ (5 points)
    \item $\{w=xy \;|\; x,y\in\Sigma^* \text{ and } \#0(x)+\#1(x)=\#0(y)+\#1(y)\}$ (5 points)
\end{enumerate}
\newpage
\subsection*{Problem 4}
As on the midterm, the shuffle operation combines two strings $x$ and $y$ in all possible
ways that preserve the original order of symbols from $x$ and $y$.  For example, if
$x=x_1x_2x_3$ and $y=y_1y_2y_3y_4$, some valid shufflings are $y_1x_1y_2x_2x_3y_3y_4$ and
$x_1x_2y_1y_2y_3x_3y_4$.  We define the shuffle of two languages $shuffle(A,B)$ to be the
shuffling of all the words in $A$ with all the words in $B$.  More explicitly,
    \[shuffle(A,B)=\bigcup\limits_{x \in A}\bigcup\limits_{y \in B} shuffle(x,y)\]
    \begin{enumerate}[(a)]
    \item Prove that, if $A$ is context-free and $B$ is regular, $shuffle(A,B)$ is
    context-free. (8 points)
    %TODO
    \item Prove that, if $A$ and $B$ are context-free, $shuffle(A,B)$ is not necessarily
    context-free. (8 points)
    %TODO
    \item For a binary language $L$ over $\Sigma=\{0,1\}$, we define the function $p$ (which
    maps a single string to a set of strings) as follows:
    \[p(w) = \{ x \in \{0,1\}^* \;|\; \#0(w) = \#0(x), \#1(w) = \#1(x) \}\]
    %TODO
    Using $p$, we define the operation $p$ on languages:

    \[p(L) = \bigcup\limits_{w \in L} p(w)\]

    Let $L$ be a regular language.  Show that $p(L)$ is context-free. (9 points)
\end{enumerate}
\newpage
\subsection*{Problem 5}

    Give a context sensitive grammar for the language:
    \[A = \{0^k \;|\; k = 2^i+1,i\ge0\}\]

\newpage
\subsection*{Problem 6}
Give a "higher level" (see "Examples of Turing machines", Sipser pp. 142) description of a
Turing machine for the language:

    \[A=\{a^ib^jc^k | i\%j=k \text{ and } i,j,k\ge1\}\]

    Where $\%$ is the modulo operator.

\end{document}
